Staged Compilation with Two-Level Type Theory

### 2022-07-28 paper/paper.tex
edit

\subsubsection{Type formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
such as functions, $\Sigma$-types, identity types or inductive types in general.
However, all constructors and eliminators in type formers must stay at the same
stage. For example:
\begin{itemize}
  \item Function domain and codomain types must be at the same stage.
  \item If we have $\Nat_0 : \U_0$ for the runtime type of natural numbers,
        we can only map from it to a type in $\U_0$ by recursion or induction.
\end{itemize}
It is not required that we have the \emph{same} type formers at both stages. We
will discuss variations of the object-level languages in Section \ref{sec:variations}.

\subsubsection{Moving between stages}
At this point, our system is rather limited, since there is no interaction
between the stages. We enable such interaction through the following operations.
Note that none of these operations can be expressed as functions, since function
types cannot cross between stages.
\begin{itemize}
\item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
  staging point of view, $\Lift A$ is the type of metaprograms which compute
  runtime expressions of type $A$.
\item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
  A quoted term $\qut{t}$ represents the metaprogram which immediately yields
  $t$.
\item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
  During staging, the metaprogram in the splice is executed, and the resulting
  expression is inserted into the output.

@@

\subsubsection{Inferring staging operations}
We can extract a coercive subtyping system from the staging operations, which
can be used for inference, and in particular for automatically transporting
definitions along preservation isomorphisms. One choice is to have $A \leq \Lift
A$, $\Lift A \leq A$, a contravariant-covariant rule for functions and a
covariant rule for $\Sigma$. During bidirectional elaboration, when we need to
compare an inferred and an expected type, we can insert coercions.  We
implemented this feature in our prototype. It additionally supports Agda-style
implicit arguments and pattern unification, so it can elaborate the following
definition:
\begin{alignat*}{3}
  & \mit{map} : \{A\,B : \Lift\U_0\} \to (\Lift A \to \Lift B)
      \to \Lift(\List_0\,A) \to \Lift(\List_0\,B)\\
  & \mit{map} := \lambda\,f\,\mit{as}.\,
      \msf{foldr}_0
        (\lambda\,a\,\mit{bs}.\, \cons_0\,(f\,a)\,\mit{bs})\,
        \nil_0\,
        as
\end{alignat*}
We may go a bit further, and also add the coercive subtyping rule $\U_0 \leq
\U_1$, witnessed by $\Lift$. Then, the type of $\mit{map}$ can be written as
$\{A\,B : \Lift\U_0\} \to (A \to B) \to \List_0\,A \to \List_0\,B$. However,
here the elaborator has to make a choice, whether to elaborate to improved or
unimproved types. In this case, the fully unimproved type would be
\[ \{A\,B : \Lift\U_0\} \to \Lift((\spl A \to \spl B) \to \List_0\,\spl A \to \List_0\,\spl B). \]
It seems that improved types are a sensible default, and we can insert explicit
lifting when we want to opt for unimproved types. This is also available in our
prototype.

@@

\subsubsection{Memory representation polymorphism}

@@

\subsubsection{Metatheory}
The metatheory is an intensional type theory with a cumulative universe
hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
1$. The transfinite $i$ levels are a convenience feature for modeling typing
contexts, where sometimes we will need universes large enough to fit types at
arbitrary finite levels. We also assume uniqueness of identity proofs and
function extensionality. We will omit transports along identity proofs in the
paper, for the sake of readability. The missing transports can be in principle
recovered due to Hofmann's conservativity result \cite{hofmann95extensional};
see also \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$,
$\top$ and $\mbb{N}$ types in all universes, and reuse Notation
\ref{notation:basics}. We also assume certain quotient inductive-inductive
types, to represent the syntax of 2LTT and the object theory. We will detail
these shortly in Sections \ref{sec:2lttmod} and \ref{sec:objmod}.

@@

\subsubsection{Elaborating informal syntax}
The justification for the usage of the informal syntax in Section
\ref{sec:tour-of-2ltt} is \emph{elaboration}. Elaboration translates surface
notation to the formal syntax (often termed ``core syntax'' in the
implementation of dependently typed programming languages). Elaboration is
partial: it may throw error on invalid input. Elaboration also requires
\emph{decidability of conversion} for the formal syntax, since type checking
requires comparing types for conversion, which in turn requires comparing terms.
We do not detail elaboration in this paper, we only make some observations.
\begin{itemize}
\item Decidability of conversion can be shown through normalization. Since 2LTT
      is a minor variation on ordinary MLTT, we expect that existing proofs for MLTT
      can be adapted without issue \cite{coquand2018canonicity,decidableconv}.
\item We need to elaborate the surface-level Russell-style universe notation to
      Coquand-style universes with explicit $\El$ and $\Code$ annotations. It is
      likely that this can be achieved with the use of bidirectional elaboration
      \cite{DBLP:journals/csur/DunfieldK21}, using two directions (checking and
      inference) both for types and terms, and inserting $\El$ and $\Code$ when
      switching between types and terms.
\item The formal complexity of elaboration varies wildly, depending on how
      explicit the surface syntax is. For example, Agda-style implicit arguments
      require metavariables and unification, which dramatically increases the
      difficulty of formalization.
\end{itemize}

\subsubsection{The setoid interpretation of the syntax}\label{sec:2lttsetoid} We need to show
that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
Unfortunately, there is no prior work which presents the setoid interpretation
of quotient types in the generality of quotient induction-induction. We sketch
it for our specific use case, but we argue that the interpretation is fairly
mechanical and could be adapted to any quotient inductive-inductive type.
We focus on contexts and types here.

@@

\subsubsection{The syntactic category and the meta-level fragment}

@@

\subsubsection{The object-level fragment}

@@

\subsubsection{Caching} In a production-strength implementation we would need some
caching mechanism, to avoid excessive duplication of code. For example, if we
use the staged $\mit{map}$ function multiple times with the same type and
function arguments, we do not want to generate separate code for each
usage. De-duplicating object-level code with function types is usually safe,
since function bodies become closed top-level code after closure conversion. We
leave this to future work.

@@

\noindent The basic scheme for internalization is as follows:
\begin{alignat*}{3}
  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
\end{alignat*}
\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
fully stable under object-theoretic substitution, so we can internalize all of
its type and term formers. We internalize object-level types and terms as
$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
together with $\NatElim$, and likewise we have all other type formers.
\subsubsection{Internal $\ev$} We also use an internal view of
$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
internalization. Below, note that the input to $\ev$ is \emph{external}, so we
mark $\ev$ as being parameterized by external input.
\begin{alignat*}{3}
  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
\end{alignat*}
\subsubsection{Object-level fragment of 2LTT}
The purely object-level syntactic fragment of 2LTT can be internalized as
follows. We define externally the presheaf of object-level types as
$|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
\Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
analogously as we have seen for $\Ty_{\mbbo j}$.

@@

\subsubsection{Preservation properties of $\re$}
First, we note that $\re$ strictly preserves $\id$, $\blank\circ\blank$ and
type/term substitution, and it preserves $\emptycon$ and $\blank\ext_1\blank$ up
to isomorphism. We have the following isomorphisms internally to $\hato$:
\begin{alignat*}{3}
  & \re_{\emptycon} &&: \re\,\emptycon \simeq \top \\
  & \re_{\ext_1}   &&: \re\,(\Gamma \ext_1 A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)
\end{alignat*}

@@

\item Meta-level positive (inductive) types are preserved in an oplax way. In the case of
  natural numbers, we have
  \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma. \]
  This is a ``serialization'' map: from a metatheoretic natural number we
  compute a numeral as a closed 2LTT term. This works analogously for other
  inductive types, including parameterized ones. For an example, from a semantic
  list of restricted terms we would get a syntactic term with list type,
  containing the same restricted terms.
\end{itemize}
\subsubsection{Action on lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
\Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
between restricted meta-level terms with type $\Lift\,A$ and restricted
object-level terms with type $A$. Hence, we write components of this isomorphism
as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the external
operations. With this, we also have $\re\,\qut{t}\,\gamma =
\qut{\re\,t\,\gamma}$ and $\re\,\spl\,t\,\gamma = \spl(\re\,t\,\gamma)$.

@@

\subsubsection{Action on object-level types and terms} $\re$ preserves all structure
in the object-level fragment of 2LTT. This follows from the $\re$ specification:
an external type $A : \Ty_{0,j}\,\Gamma$ is directly internalized as an element
of $\Ty_{0,j}$, and the same happens for $t : \Tm_{0,j}\,\Gamma\,A$.

@@

\subsubsection{Syntactic category and terminal object}
Here, we simply have $\id\rel\,\gamma\rel := \gamma\rel$ and $(\sigma \circ
\delta)\rel\,\gamma\rel := \sigma\rel\,(\delta\rel\,\gamma\rel)$. The terminal
object is interpreted as $\emptycon\rel\,\gamma\,\gamma' := \top$.

\subsubsection{Meta-level family structure}
We interpret context extension and type/term substitution as follows. Note the
usage of the pattern matching notation on the $\rexti$ isomorphism.
\begin{alignat*}{4}
  & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
    (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
  & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
  & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
\end{alignat*}
It is enough to specify the $\blank\rel$ action on extended substitutions
$(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
\ext A)\,(A[\p])$. The category-with-families equations hold evidently.
\begin{alignat*}{4}
  & (\sigma,\,t)\rel\,\gamma\rel := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \hspace{2em} \p\rel\,(\gamma\rel,\,t\rel) := \gamma\rel \hspace{2em} \q\rel\,(\gamma\rel,\,t\rel) := t\rel
\end{alignat*}

\subsubsection{Meta-level natural numbers} First, we have to define a relation:
\begin{alignat*}{4}
  & (\Nat_{1,j})\rel : \Gamma\rel\,\gamma\,\gamma' \to \mbb{N} \to \re\,\Nat_{1,j}\,\gamma' \to \whset_j\\
  & (\Nat_{1,j})\rel\,\gamma\rel\,n\,n' := (\re_{\mbb{N}}\,n = n')
\end{alignat*}
Note that evaluation sends $\Nat_{1,j}$ to the semantic type of natural numbers,
i.e.\ $\ev\,\Nat_{1,j} = \mbb{N}$. We refer to the serialization map
$\re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma'$ above. In short,
$(\Nat_{1,j})\rel$ expresses \emph{canonicity}: $n'$ is canonical precisely if
it is of the form $\re_{\mbb{N}}\,n$ for some $n$.

@@

\subsubsection{Meta-level $\Sigma$-types} We define relatedness pointwise. Pairing and projection
are interpreted as meta-level pairing and projection.
\begin{alignat*}{3}
  &(\Sigma\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \times \ev\,B\,(\gamma,\,\alpha))
             \to \re\,(\Sigma\,A\,B)\,\gamma' \to \whset_j\\
  &(\Sigma\,A\,B)\rel\,\gamma\rel\,(\alpha,\,\beta)\,(\re_{\Sigma}^{-1}(\alpha',\,\beta')) :=
             (\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \times B\rel\,(\gamma\rel,\,\alpha\rel)\,\beta\,\beta'
\end{alignat*}

\subsubsection{Meta-level $\Pi$-types} We again use a pointwise definition. Note that we need to use
$\re_{\app}$ to apply $t' : \re\,(\Pi\,A\,B)\,\gamma'$ to $\alpha'$.
\begin{alignat*}{3}
  &(\Pi\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \to \ev\,B\,(\gamma,\,\alpha))
             \to \re\,(\Pi\,A\,B)\,\gamma' \to \whset_j\\
  &(\Pi\,A\,B)\rel\,\gamma\rel\,t\,t' :=
    (\alpha : \ev\,A\,\gamma)(\alpha' : \re\,A\,\gamma')(\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \to
    B\rel\,(\gamma\rel,\,\alpha\rel)\,(t\,\alpha)\,(\re_{\app}\,t'\,\alpha')
\end{alignat*}
For abstraction and application, we use a curry-uncurry definition:
\begin{alignat*}{4}
  &(\lam\,t)\rel\,\gamma\rel                &&:= \lambda\,\alpha\,\alpha'\,\alpha\rel.\,t\rel\,(\gamma\rel,\,\alpha\rel)\\
  &(\app\,t)\rel\,(\gamma\rel,\,\alpha\rel) &&:= t\rel\,\gamma\rel\,\alpha\,\alpha'\,\alpha\rel
\end{alignat*}

>>>>

\subsubsection{Type Formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
such as functions, $\Sigma$-types, identity types or inductive types in general.
However, all constructors and eliminators in type formers must stay at the same
stage. For example:
\begin{itemize}
  \item Function domain and codomain types must be at the same stage.
  \item If we have $\Nat_0 : \U_0$ for the runtime type of natural numbers,
        we can only map from it to a type in $\U_0$ by recursion or induction.
\end{itemize}
It is not required that we have the \emph{same} type formers at both stages. We
will discuss variations of the object-level languages in Section \ref{sec:variations}.

\subsubsection{Moving Between Stages}
At this point, our system is rather limited, since there is no interaction
between the stages. We enable such interaction through the following operations.
Note that none of these operations can be expressed as functions, since function
types cannot cross between stages.
\begin{itemize}
\item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
  staging point of view, $\Lift A$ is the type of metaprograms which compute
  runtime expressions of type $A$.
\item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
  A quoted term $\qut{t}$ represents the metaprogram which immediately yields
  $t$.
\item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
  During staging, the metaprogram in the splice is executed, and the resulting
  expression is inserted into the output.

@@

\subsubsection{Inferring Staging Operations}
We can extract a coercive subtyping system from the staging operations, which
can be used for inference, and in particular for automatically transporting
definitions along preservation isomorphisms. One choice is to have $A \leq \Lift
A$, $\Lift A \leq A$, a contravariant-covariant rule for functions and a
covariant rule for $\Sigma$. During bidirectional elaboration, when we need to
compare an inferred and an expected type, we can insert coercions.  We
implemented this feature in our prototype. It additionally supports Agda-style
implicit arguments and pattern unification, so it can elaborate the following
definition:
\begin{alignat*}{3}
  & \mit{map} : \{A\,B : \Lift\U_0\} \to (\Lift A \to \Lift B)
      \to \Lift(\List_0\,A) \to \Lift(\List_0\,B)\\
  & \mit{map} := \lambda\,f\,\mit{as}.\,
      \msf{foldr}_0
        (\lambda\,a\,\mit{bs}.\, \cons_0\,(f\,a)\,\mit{bs})\,
        \nil_0\,
        as
\end{alignat*}
We may go a bit further, and also add the coercive subtyping rule $\U_0 \leq
\U_1$, witnessed by $\Lift$. Then, the type of $\mit{map}$ can be written as
$\{A\,B : \Lift\U_0\} \to (A \to B) \to \List_0\,A \to \List_0\,B$. However,
here the elaborator has to make a choice, whether to elaborate to improved or
unimproved types. In this case, the fully unimproved type would be
\[ \{A\,B : \Lift\U_0\} \to \Lift((\spl A \to \spl B) \to \List_0\,\spl A \to \List_0\,\spl B). \]
It seems that improved types are a sensible default, and we can insert explicit
lifting when we want to opt for unimproved types. This is also available in our
prototype.

@@

\subsubsection{Memory Representation Polymorphism}

@@

\paragraph{Metatheory}
The metatheory is an intensional type theory with a cumulative universe
hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
1$. The transfinite $i$ levels are a convenience feature for modeling typing
contexts, where sometimes we will need universes large enough to fit types at
arbitrary finite levels. We also assume uniqueness of identity proofs and
function extensionality. We will omit transports along identity proofs in the
paper, for the sake of readability. The missing transports can be in principle
recovered due to Hofmann's conservativity result \cite{hofmann95extensional};
see also \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$,
$\top$ and $\mbb{N}$ types in all universes, and reuse Notation
\ref{notation:basics}. We also assume certain quotient inductive-inductive
types, to represent the syntax of 2LTT and the object theory. We will detail
these shortly in Sections \ref{sec:2lttmod} and \ref{sec:objmod}.

@@

\subsubsection{Elaborating Informal Syntax}
The justification for the usage of the informal syntax in Section
\ref{sec:tour-of-2ltt} is \emph{elaboration}. Elaboration translates surface
notation to the formal syntax (often termed ``core syntax'' in the
implementation of dependently typed programming languages). Elaboration is
partial: it may throw error on invalid input. Elaboration also requires
\emph{decidability of conversion} for the formal syntax, since type checking
requires comparing types for conversion, which in turn requires comparing terms.
We do not detail elaboration in this paper, we only make some observations.
\begin{itemize}
\item Decidability of conversion can be shown through normalization. Since 2LTT
      is a minor variation on ordinary MLTT, we expect that existing proofs for MLTT
      can be adapted without issue \cite{coquand2018canonicity,decidableconv}.
\item We need to elaborate the surface-level Russell-style universe notation to
      Coquand-style universes with explicit $\El$ and $\Code$ annotations. It is
      likely that this can be achieved with the use of bidirectional elaboration
      \cite{DBLP:journals/csur/DunfieldK21}, using two directions (checking and
      inference) both for types and terms, and inserting $\El$ and $\Code$ when
      switching between types and terms.
\item The formal complexity of elaboration varies wildly, depending on how
      explicit the surface syntax is. For example, Agda-style implicit arguments
      require metavariables and unification, which dramatically increases the
      difficulty of formalization.
\end{itemize}

\subsubsection{The Setoid Interpretation of the Syntax}\label{sec:2lttsetoid} We need to show
that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
Unfortunately, there is no prior work which presents the setoid interpretation
of quotient types in the generality of quotient induction-induction. We sketch
it for our specific use case, but we argue that the interpretation is fairly
mechanical and could be adapted to any quotient inductive-inductive type.
We focus on contexts and types here.

@@

\subsubsection{The Syntactic Category and the Meta-Level Fragment}

@@

\subsubsection{The Object-Level Fragment}

@@

\paragraph{Caching} In a production-strength implementation we would need some
caching mechanism, to avoid excessive duplication of code. For example, if we
use the staged $\mit{map}$ function multiple times with the same type and
function arguments, we do not want to generate separate code for each
usage. De-duplicating object-level code with function types is usually safe,
since function bodies become closed top-level code after closure conversion. We
leave this to future work.

@@

\noindent The basic scheme for internalization is as follows:
\begin{alignat*}{3}
  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
\end{alignat*}
\subsubsection{Object-Theoretic Syntax} The syntax of the object theory is clearly
fully stable under object-theoretic substitution, so we can internalize all of
its type and term formers. We internalize object-level types and terms as
$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
together with $\NatElim$, and likewise we have all other type formers.
\subsubsection{Internal $\ev$} We also use an internal view of
$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
internalization. Below, note that the input to $\ev$ is \emph{external}, so we
mark $\ev$ as being parameterized by external input.
\begin{alignat*}{3}
  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
\end{alignat*}
\subsubsection{Object-Level Fragment of 2LTT}
The purely object-level syntactic fragment of 2LTT can be internalized as
follows. We define externally the presheaf of object-level types as
$|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
\Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
analogously as we have seen for $\Ty_{\mbbo j}$.

@@

\subsubsection{Preservation Properties of $\re$}
First, we note that $\re$ strictly preserves $\id$, $\blank\circ\blank$ and
type/term substitution, and it preserves $\emptycon$ and $\blank\ext_1\blank$ up
to isomorphism. We have the following isomorphisms internally to $\hato$:
\begin{alignat*}{3}
  & \re_{\emptycon} &&: \re\,\emptycon \simeq \top \\
  & \re_{\ext_1}   &&: \re\,(\Gamma \ext_1 A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)
\end{alignat*}

@@

\item Meta-level positive (inductive) types are preserved in an oplax way. In the case of
  natural numbers, we have
  \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma. \]
  This is a ``serialization'' map: from a metatheoretic natural number we
  compute a numeral as a closed 2LTT term. This works analogously for other
  inductive types, including parameterized ones. For an example, from a semantic
  list of restricted terms we would get a syntactic term with list type,
  containing the same restricted terms.
\end{itemize}
\subsubsection{Action on Lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
\Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
between restricted meta-level terms with type $\Lift\,A$ and restricted
object-level terms with type $A$. Hence, we write components of this isomorphism
as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the external
operations. With this, we also have $\re\,\qut{t}\,\gamma =
\qut{\re\,t\,\gamma}$ and $\re\,\spl\,t\,\gamma = \spl(\re\,t\,\gamma)$.

@@

\subsubsection{Action on Object-Level Types and Terms} $\re$ preserves all structure
in the object-level fragment of 2LTT. This follows from the $\re$ specification:
an external type $A : \Ty_{0,j}\,\Gamma$ is directly internalized as an element
of $\Ty_{0,j}$, and the same happens for $t : \Tm_{0,j}\,\Gamma\,A$.

@@

\subsubsection{Syntactic Category and Terminal Object}
Here, we simply have $\id\rel\,\gamma\rel := \gamma\rel$ and $(\sigma \circ
\delta)\rel\,\gamma\rel := \sigma\rel\,(\delta\rel\,\gamma\rel)$. The terminal
object is interpreted as $\emptycon\rel\,\gamma\,\gamma' := \top$.

\subsubsection{Meta-Level Family Structure}
We interpret context extension and type/term substitution as follows. Note the
usage of the pattern matching notation on the $\rexti$ isomorphism.
\begin{alignat*}{4}
  & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
    (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
  & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
  & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
\end{alignat*}
It is enough to specify the $\blank\rel$ action on extended substitutions
$(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
\ext A)\,(A[\p])$. The category-with-families equations hold evidently.
\begin{alignat*}{4}
  & (\sigma,\,t)\rel\,\gamma\rel := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \hspace{2em} \p\rel\,(\gamma\rel,\,t\rel) := \gamma\rel \hspace{2em} \q\rel\,(\gamma\rel,\,t\rel) := t\rel
\end{alignat*}

\subsubsection{Meta-Level Natural Numbers} First, we have to define a relation:
\begin{alignat*}{4}
  & (\Nat_{1,j})\rel : \Gamma\rel\,\gamma\,\gamma' \to \mbb{N} \to \re\,\Nat_{1,j}\,\gamma' \to \whset_j\\
  & (\Nat_{1,j})\rel\,\gamma\rel\,n\,n' := (\re_{\mbb{N}}\,n = n')
\end{alignat*}
Note that evaluation sends $\Nat_{1,j}$ to the semantic type of natural numbers,
i.e.\ $\ev\,\Nat_{1,j} = \mbb{N}$. We refer to the serialization map
$\re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma'$ above. In short,
$(\Nat_{1,j})\rel$ expresses \emph{canonicity}: $n'$ is canonical precisely if
it is of the form $\re_{\mbb{N}}\,n$ for some $n$.

@@

\subsubsection{Meta-Level $\Sigma$-Types} We define relatedness pointwise. Pairing and projection
are interpreted as meta-level pairing and projection.
\begin{alignat*}{3}
  &(\Sigma\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \times \ev\,B\,(\gamma,\,\alpha))
             \to \re\,(\Sigma\,A\,B)\,\gamma' \to \whset_j\\
  &(\Sigma\,A\,B)\rel\,\gamma\rel\,(\alpha,\,\beta)\,(\re_{\Sigma}^{-1}(\alpha',\,\beta')) :=
             (\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \times B\rel\,(\gamma\rel,\,\alpha\rel)\,\beta\,\beta'
\end{alignat*}

\subsubsection{Meta-Level $\Pi$-Types} We again use a pointwise definition. Note that we need to use
$\re_{\app}$ to apply $t' : \re\,(\Pi\,A\,B)\,\gamma'$ to $\alpha'$.
\begin{alignat*}{3}
  &(\Pi\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \to \ev\,B\,(\gamma,\,\alpha))
             \to \re\,(\Pi\,A\,B)\,\gamma' \to \whset_j\\
  &(\Pi\,A\,B)\rel\,\gamma\rel\,t\,t' :=
    (\alpha : \ev\,A\,\gamma)(\alpha' : \re\,A\,\gamma')(\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \to
    B\rel\,(\gamma\rel,\,\alpha\rel)\,(t\,\alpha)\,(\re_{\app}\,t'\,\alpha')
\end{alignat*}
For abstraction and application, we use a curry-uncurry definition:
\begin{alignat*}{4}
  &(\lam\,t)\rel\,\gamma\rel                &&:= \lambda\,\alpha\,\alpha'\,\alpha\rel.\,t\rel\,(\gamma\rel,\,\alpha\rel)\\
  &(\app\,t)\rel\,(\gamma\rel,\,\alpha\rel) &&:= t\rel\,\gamma\rel\,\alpha\,\alpha'\,\alpha\rel
\end{alignat*}

\subsubsection{Meta-Level Universes} We interpret $\U_{1,j}$ as a semantic relation space:
\begin{alignat*}{3}
  &(\U_{1,j})\rel\,\gamma\rel : \whset_j \to \re\,\U_{1,j}\,\gamma' \to \whset_{j+1} \\
  &(\U_{1,j})\rel\,\gamma\rel\,t\,t' := t \to \re_{\El}\,t' \to \whset_{j}
\end{alignat*}
Note that we have
\begin{alignat*}{3}
  & (\El\,t)\rel   &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j\\
  & (\Code\,t)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j.
\end{alignat*}
The types coincide because of the equation $\re\,(\El\,t)\,\gamma' =
\re_{\El}\,(\re\,t\,\gamma')$. Therefore we can interpret $\El$ and $\Code$ as
identity maps, as $(\El\,t)\rel := t\rel$ and $(\Code\,t)\rel := t\rel$.
